\documentclass{article}
\usepackage{algpseudocode}
\usepackage{amsmath}
\usepackage{latexsym}
\usepackage[super]{nth}
\title{MeshAdapt}
\author{Dan Ibanez}
\begin{document}

\maketitle

\section{Coarsening}

Garimella outlines conditions for an
edge collapse to be valid, which have since then been refined
into the following four conditions:

\begin{enumerate}
\item If the vertices are classified on equal order model entities,
then both vertices and the edge must have the same classification.
\item If the vertices are classified on different orders, then
the vertex to be collapsed and the edge must be classified on
the higher-order entity.
\item If there exists a ring of three edges containing the
collapsing edge, there must exist a face bounded by this ring.
The non-collapsing edge adjacent to the collapsing vertex
must have the same classification as this face.
\item If there exist two edge-adjacent faces that connect
the vertices, those faces and the edge must bound a tet.
The face adjacent to the collapsing vertex must
have the same classification as this tet.
\end{enumerate}

The last two conditions can be checked by identifying the generalized
"ring" cases and using findUpward in the 2D case or a specialized
search on the adjacent regions of a face in the 3D case.

For condition 4, here referred to as the face ring condition,
it is useful to optimize the procedure since a naive implementation
compares adjacent faces of one vertex to another, performing $n^2$
comparisons where $n$ is the number of faces adjacent to a vertex.
$n$ is on average 30 and can be as high as 78 as measured in some tet
meshes, and this has been measured as the bottleneck in the current
coarsening procedure.

De Cougny outlines an algorithm for identifying the independent set
of vertices to not collapse in order to maintain a decent vertex
distribution:
\begin{algorithmic}
\For{$d \in [0..n_{sd}]$}
\For{$v \sqsubset G^d_j$}
\State $v \in S$
\For{$e \in v\{M^1\}$}
\If{$e \in E_{\text{collapse}}$}
\State $v_2 = (e\{M^0\})\setminus v$
\If{$v_2\in S$}
\If{$(v\sqsubset g)\land(v_2\sqsubset\bar{g})$}
\State $v\not\in S$
\EndIf
\EndIf
\EndIf
\EndFor
\EndFor
\EndFor
\end{algorithmic}
Basically, given a target vertex, if another vertex is adjacent via
a collapsing edge and the other vertex is classified on the closure
of the classification of the target, then the target is not placed
in the set.

Which is just another way of stating the definition of vertices to
collapse as defined by conditions 1 and 2
described by Garimella.

In fact, we can use those conditions to more efficiently compute the
set by avoiding the check $v_2\sqsubset\bar{g}$, which requires checking
for classification on the closure of a model entity $g$, which may be
expensive (model adjacency counts are not bounded in any way).

If we first reverse the problem and choose to only identify vertices
not in the independent set $v\in\bar{S}$ (i.e. vertices which we intend to collapse),
then the algorithm can proceed as follows:
\begin{enumerate}

\item All vertices are implicitly assumed to be in $S$, $\bar{S}$ is empty.

\item Identify candidate edges and vertices for collapsing based on the initial
criteria (a mesh size field) and Garimella's first 2 conditions.
If the edge satisfies condition 1, both vertices have equal classification and are on the closures
of one another's classifications, and both are placed in $\bar{S}$ tentatively.
If the edge satisfies condition 2, only the vertex with higher order classification
is a candidate and the non-candidate is clearly on the closure of the candidate.
The candidate is placed in $\bar{S}$.

\item Run the modified graph algorithm as follows:
\begin{algorithmic}
\For{$d \in [0..n_{sd}]$}
\For{$\{v|v\in\bar{S}, \sqsubset G^d_j\}$}
\State {remove $v$ from $\bar{S}$}
\For{$e \in v\{M^1\}$}
\If{$e \in E_{\text{collapse}}$}
\State $v_2 = (e\{M^0\})\setminus v$
\If{$v_2\not\in\bar{S}$}
\State {put $v$ back in $\bar{S}$}
\EndIf
\EndIf
\EndFor
\EndFor
\EndFor
\end{algorithmic}
\end{enumerate}

Note that the independent set algorithm can be most simply expressed as:
mark a vertex as illegal to collapse unless doing so would prevent a marked
edge from collapsing.
By the end of the procedure, every vertex which is legal to collapse is required
for some particular edge to collapse.

In order to prevent confusion about which edge collapse is associated with a vertex
to collapse, we coarsen one geometric dimension at a time, starting with model edges.
The topological, classification, and independent set routines operate only over
edges and vertices classified on model entities of dimension $d$, where $d$ increases from 1.

Garimella's conditions 3 and 4 must be checked right before executing the collapse,
since nearby edge collapses can affect the outcome of these topological checks.

Collapsing an edge involves two sets of elements: the set of elements to collapse
and the set of elements to keep.
The set of elements to collapse are those adjacent to the edge to collapse,
and the set of elements to keep are those adjacent to the vertex to collapse
but not to the edge to collapse.
To collapse an edge, the elements to keep are first reconstructed by with
the vertex to collapse replaced by the vertex to keep.
Care must be taken to exactly duplicate the geometric classification of
entities as they are reconstructed.
After that reconstruction, the elements to collapse are removed, taking
with them the edge and vertex to collapse.

Before an edge is collapsed, a final check is performed to ensure no invalid
elements will be created.
Since the mesh is straight-sided, the check temporarily moves the vertex to
collapse onto the vertex to keep and then checks the resulting volumes of the elements to keep.

All three steps of coarsening: initial checks, independent set finding, and collapsing
operate on a single geometric model dimension using the cavity operator
framework provided in APF.
For the initial checks, the cavity includes all elements adjacent to both
vertices of the edge.
For independent set, the cavity is the elements around a vertex.
Finally, an edge collapse requires all elements adjacent to the vertex to collapse.

\section{Refinement}

A single step of refinement is run as follows:
\begin{enumerate}
\item select edges to split
\item for each dimension, split entities of that dimension
by just creating new entities,
classifying those new entities the same as the parent being split
and performing solution transfer.
\item link the edge split vertices based on the remote
copies of their parent edges.
\item invoke the refinement templates on all mesh entities,
one dimension at a time from edges up to regions.
\item remove all old entities that have been split
\item link all other remote copies based on the remote copies
of vertices
\end{enumerate}

The majority of the complexity in refinement is the use of
whole-entity refinement templates as opposed to individual
edge splits.
Refinement is well-defined for simplex meshes, and
templates exist for triangular faces and tetrahedra.
In general, for a simplex, there is one template for
each rotationally unique combination of split/unsplit edges.
The total number of combinations is $2^e$ for a simplex
with $e$ edges.
The number of rotationally unique combinations is smaller:
3 for triangles and 12 for tetrahedra.

We first define lookup tables from the full set of combinations
into the rotationally unique ones.
To form a lookup table, we encode a combination as an integer
in $[0..2^e-1]$ using binary numbers where each bit represents
the state of one edge.
These are called edge codes.
A set of codes representing the rotationally unique combinations
is chosen as the set of canonical edge codes.
The lookup table maps all edge codes to a rotation and a
canonical edge code.

Rotation itself is a key tool for implementing refinement templates,
and rotation functions and their properties are at the heart of
the system.

Also at the center of the code's simplicity is the representation
of all domains as a set of vertices, and manipulating only
vertex sets.
This is supported by a set of entity creation functions that
accept vertex sets and recursively construct the closure
of the entity, using existing entities if possible.

Given that all domains are represented by a set of vertices,
we need to know the geometry of each domain and provide
rotation functions for the geometries of interest.

The usefulness of all this becomes apparent when templates
are decomposed into sub-problems, where a sub-problem may
involve tetrahedronizing a subspace of the parent tetrahedron
that is shaped like a pyramid or a prism or an octahedron
whose bounding vertices are known.
This delegation to sub-problems shortens the code, which
greatly improves maintainability and reduces the probability
of programming errors.

It also allows more advanced decisions, such as choosing
the shortest edge when triangulating a quad or an octahedron,
to be represented by one function for all cases.

\section{Size Field}

The size field is a \nth{2}-order tensor field over the mesh.
It is meant to indicate the anisotropic desired fineness or
coarseness of discretization along a particular direction
at any point in space.

One way to interpret the size field is as a generalization of the dot product.
Where a vector $v$ at a particular point $p$ in space would have
been measured as $v\cdot v$ (not dependent on $p$), it is now
measured using the metric tensor $T(p)$ at $p$: $vT(p)v^T$.

Another interpretation is to decompose the metric tensor into
the product of a transformation matrix times itself: $T=QQ^T$.
One can then see that the metric measures a transformed vector
$v'=vQ$: $vTv^T=vQQ^Tv^T=v'(v')^T=v'\cdot v'$.

The transformation $Q$ can be further decomposed into a rotation $R$
and scaling $S$:
\[Q=RS,
R=\begin{bmatrix}
e_1 & e_2 & e_3
\end{bmatrix},
S=\begin{bmatrix}
\frac{1}{h_0} & 0 & 0 \\
0 & \frac{1}{h_1} & 0 \\
0 & 0 & \frac{1}{h_2} \\
\end{bmatrix}\]
Where $e_1,e_2,e_3$ are the principal axes of the transformed
space expressed as real space unit vectors and
$h_0,h_1,h_2$ are the desired edge lengths along each
principal axis.

The main complexity with this definition is that we have a non-trivial
metric tensor that varies depending on position in real space, which
makes measuring objects in real space using this metric field somewhat
difficult.

Algorithms in MeshAdapt are mainly interested in the length of edges
in metric space and the volume of elements in metric space.
There exists a generalized definition of volume called a measure,
which is currently computed as follows:

\[V=\int_{\Omega^e} |J(\xi)|\]

In particular, $|J(\xi)|=d\Omega^e$ is the determinant of the Jacobian $J=x_{,\xi}$.
Since we are interested in measuring edges (1D elements embedded in 3D)
and faces (2D elements embedded in 3D), then $J$ sometimes has zero rows
and $|J|$ is a more specialized computation that takes into account
the dimension of the element.
Regardless of dimension, $J$ represents the differential tangent vectors
along the element's local coordinates, and transforming them into the
metric space before taking the determinant is an simple way to obtain
differential volume in the metric space:

\[V'=\int_{\Omega^e} |J(\xi)Q(\xi)|\]

Both real and metric space measures are computed using Gaussian quadrature:

\[V'=\sum_{l=1}^{\text{nint}} |J(\xi_l)Q(\xi_l)|\]

If the metric field is encoded as a \nth{2}-order tensor $Q(x)$ distributed
over the mesh as a nodal field, then the existing framework for interpolation
of $Q$ and performing quadrature are leveraged to easily measure any
mesh entity in the metric space.

\end{document}
