\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage[super]{nth}
\title{Interface for Field Access}
\author{Dan Ibanez, SCOREC}
\begin{document}
\maketitle

\section{Goal}

This document aims to establish a consensus on
basic operations that should be made available
by finite element toolkits for accessing finite
element field data attached to a mesh.

\section{Definitions}

A finite element mesh $M$ represents a domain
$\Omega^h \subset {\mathbb R^D}$
composed of entity domains $\Omega_e$:
\[\Omega^h = \bigcup_{e\in M} \Omega_e\]

A finite element field $f : \Omega^h \to T$ is a piece-wise field
from the mesh domain to some tensor space $T$, where
each piece is $f_e : \Omega_e \to T$.
The entity domains have parametric, or parent, coordinate
systems and a map from parametric points $\xi \in {\mathbb R^d}$
to points in the entity domain $X_e(\xi) \in \Omega_e$.

The value of a field at a point $x$ in an entity domain $\Omega_e$
is entirely defined by the corresponding parametric point $\xi$
and a set of tensor multipliers $\bar{W}_e = \{\bar{W}_e^0,...,\bar{W}_e^n\}$
which affect that entity:
\[x\in \Omega_e, x = X_e(\xi), f(x) = f_e(\xi,\bar{W}_e)\]
For example, in a linear tetrahedral mesh the set $\bar{W}_e$ contains
the field values at the vertices of an entity.

The field multipliers are each associated with a unique mesh entity
to allow sharing between entities.
The set of multiplier tensors associated with entity $e$ is denoted
$W_e = \{W_e^0,...,W_e^m\}$.
The set of tensors which affect the field definition over an entity
is composed of tensors associated to nearby entities, usually to
entities in its closure:
\[\bar{W}_e = \bigcup_{a \in \bar{e}} W_a\]
To reuse our example, in linear tetrahedral meshes multipliers are
the field values associated with vertices, and the multipliers
which affect a tetrahedron are those associated to its vertices.

\section{Requirements}

To describe a field, two key pieces of information are needed:
\begin{enumerate}
\item The entity basis functions $f_e(\xi,\bar{W}_e)$ for all $e\in M$.
\item The associated multipliers $W_e$ for all $e \in M$
\end{enumerate}
The description of $f_e$ encompasses several smaller pieces of information:
\begin{enumerate}
\item The parent domain $\Box_e$ such that $X_e(\Box_e)=\Omega_e$.
\item The how many multipliers are associated with each entity: $|W_e|$.
\item How to collect all affecting multipliers: $g(e) = \bar{W}_e$.
\end{enumerate}
Depending on the complexity of the field, much of this information
can be recorded {\it a priori}.
For example, the description of $f_e$ may be just ``\nth{2}-order Lagrange"
or ``Cubic B\'ezier".
In more complex cases such as variable-order fields, information
such as $|W_e|$ and $g$ may have to be specified on a per-entity basis.

\section{Coordinates}

Note that the coordinates are themselves a field $X_e : \Box_e \to \Omega_e$,
and should have equivalent availability of information.
For example, it should be possible to query the basis functions
being used to define coordinates.

\section{Interface}

We will assume an adequate existing interface for topological
mesh descriptions, so that operations such as iterating over
$e \in M$ and getting the entities $a \in \bar{e}$ are available
and efficient.

At a high level, it should also be possible to query the set of fields
defined over a particular mesh.
Fields and entities should have identifying handles that can be efficiently
interpreted by the software.

At a minimum, just to preserve field data, it must be possible to access
tensor multipliers associated with entities.
That is, for a given entity $e$, the count $n=|W_e|$ and values
$W_e^i,1\leq i\leq n$ must be quickly accessible.
Of course, prerequisite to retrieving $W_e^i$ is some agreement on the
storage format for tensor values.
For each $f : \Omega^h \to T$,
the storage format for $t\in T$ should be accessible.

To preserve a complete description of the field, it should also be
possible to query the exact basis functions $f_e$ being used for
each entity, along with the closely associated procedure $g$ for
collecting the ordered set of tensor multipliers $\bar{W}_e$ that 
are weighted by $f_e$ based on $\xi$ to produce $f(x)$.

\end{document}
