\documentclass{article}
\usepackage{algpseudocode}
\title{The Cavity Operator}
\author{Dan Ibanez}
\begin{document}

\maketitle

\section{Overview}

The Cavity Operator system, or CavityOp, is a framework
for many of the operations carried out in field computation
and mesh adaptation.
The fundamental characteristic of a cavity operator is that
it queries and modifies a local cavity around a given mesh
entity.
The CavityOp system exists to allow parallel applications to
write their cavity-based logic in a serial mindset.
The CavityOp runtime system then uses migration to provide
local cavities to the user's operator for all entities.

\section{User Operator}

The operator provided by the user should define routines for
the following:
\begin{enumerate}
\item Determining whether the operator should apply to an entity
\item Building the cavity based on adjacencies while
marking which entities need to be local
\item Applying the operator to a cavity
\end{enumerate}
It should also follow these rules:
\begin{enumerate}
\item Building the cavity should not modify anything
\item Only local entities should be modified, and only cavity
entities should be queried.
\item If upward adjacencies of an entity are added to the cavity,
that entity must first be local
\end{enumerate}
A mesh entity is local when it has no remote copies.
Given other assumptions about the mesh, this also means
that all its upward adjacencies are fully retrievable.

The user then gives this operator object containing the definitions
to a routine that will apply the operator to every mesh
entity of a given dimension.
So far all operators target a particular dimension, but any
iteration over the mesh fits this algorithm.
The job of the CavityOp application routine is to use migration
to deal with cavities that would otherwise be split by a
partition boundary.

\section{User Example}

To illustrate what a cavity operator would do, we can use the simple
example of averaging per-element values into per-vertex values.
In an error estimator, each element receives a scalar target size,
and the size field value at a vertex is the average of its adjacent
elements' target sizes.

A user operator for this would define the three functions described
earlier.
To determine whether it should apply to a vertex, it would check
whether the size value has already been computed and attached to the vertex.
To build the cavity, it must first insist that the vertex be local,
then gather the upward adjacent elements of the vertex.
To apply the operator, it queries the element values and attaches
their average to the vertex.

Notice that the vertex's locality is enforced before querying upward
adjacencies or modifying its field data.

\section{Pull Requests}

The first concept towards understanding CavityOp is the pull request.
This is a part requesting that a mesh entity be made local to it.
In order to make an entity local, all its upward adjacent elements
must be on the same part.
So a pull request is really asking for all adjacent elements to
move to the requesting part.
Loosely, the cavity operator system is just trying to satisfy pull requests
until there are no more.
Given an algorithm that is guaranteed to make some progress on those pull requests,
the application function will repeat these steps until completion:
\begin{enumerate}
\item apply the operator to all valid local cavities
\item identify paritioned cavities and submit pull requests
\item try to satisfy pull requests
\end{enumerate}
More specifically, here is pseudo-code for the application of a cavity operator:

\begin{algorithmic}
\Function{apply}{$d,o$}
\State let $d$ be the dimension of target entities
\State let $o$ be the operator
\Loop
  \For {$e \in M\{M^d\}$ of the local part}
    \If {$e$ is owned and $o$ should apply to $e$}
      \State try to build cavity around $e$
      \If {locality requirements were met}
        \State apply $o$ to $e$
      \EndIf
    \EndIf
  \EndFor
  \For {$e \in M\{M^d\}$ of the local part}
    \If {$e$ is owned and $o$ should apply to $e$}
      \State try to build cavity around $e$
      \If {some entities that need to be local are not}
        \State submit pull requests for them
      \EndIf
    \EndIf
  \EndFor
  \If {no pull requests were submitted}
    \Return
  \EndIf
  \State try to satisfy pull requests
\EndLoop
\EndFunction
\end{algorithmic}

\section{Satisfying Pull Requests}

A pull request for an entity is satisfied if all the
entity's adjacent elements are migrated to one part.
This naturally creates conflicts when two or more pull requests
try to move an element to different parts.

\emph{When this happens, the element is migrated to the part
with the largest integer ID.}

As long as this conflict resolution rule is a strict
ordering of the parts, it guarantees that at least
one pull request is satisfied during migration.
In fact, the part with the highest ID that submits
pull requests will have them all satisfied, since
they win all conflicts.

For an operator that will be applied to a fixed
number of mesh entities, as is so far the case,
this is guaranteed to terminate since
targets are always eliminated after each migration.

\section{Performance}

The performance of this algorithm depends on
the ability to satisfy pull requests.
In the end, the problem can be reduced
to finding the maximal independent set
of a graph where the nodes are cavities 
and the edges are pull request conflicts.
In practice, the resolution heuristic should give
a decent independent set.

This is helped by the fact that operator application
is based on ownership, and the PUMI mesh database
selects ownership over partition model entities,
so there are rarely conflicts in the middle of a
partition model face.

For varying cavities such as SPR or mesh-modifying
operators, it is even harder to prove anything.

The expectation is that the number of migrations should
be a small constant not dependent on mesh size
or part count.
Early testing on a 10K region SLAC mesh used 2 migrations
to completely perform SPR for 2 parallel parts and 3 migrations
for 4 parallel parts.
More performance data is now being gathered and
should be available within a few days.

\end{document}
