\documentclass{article}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{algpseudocode}
\title{Parallel Stitching}
\author{Dan Ibanez}
\begin{document}
\maketitle

\section{Theory}

This section outlines the theorical proof of the algorithm
to follow which sets up a partition model and remote
copy links from incomplete data for a distributed mesh.
It assumes a full one-level mesh representation with
correct partition model classification.

In a distributed mesh, each mesh entity of the entire mesh
resides on a set of parts.
If a mesh entity resides on a part, then a copy of it exists
in the mesh data structure of that part.
Let $\sim$ denote residence: $e\sim p$ means mesh entity $e$ resides
on part $p$.
\[(e \sim p)\equiv(e \sqsubset \bar{P^3_p})\]
Along the same lines, let $R(e)$ denote the resident part set of
a mesh entity, which consist of all parts on which the entity resides.
\[R(e) = \{p|(e\sim p)\}\]
Also let $\downarrow$ denote the one-level downward adjacency of
a mesh entity.
\[\downarrow e = e\{M^{d-1}\}, e \in M^d\]

A fundamental axiom of distributed meshes is this: if a
mesh entity resides on a part, its boundary must also reside
on that part.
\[(e\sim p) \to (\partial e\sim p)\]
The same goes for the downward adjacent entities, since they are contained
in the boundary.
\[\downarrow e \subseteq \partial e,
(e\sim p)\to(\downarrow e\sim p)\]
Therefore, if the downward entities do
not fully reside on a part, the entity cannot reside on that part.
\[\neg (\downarrow e\sim p) \to \neg (e\sim p)\]

Given this, we can define the set of parts on which a mesh entity
is allowed to reside as the intersection of the resident sets of
its downward entities, since these are the parts on which
they all reside.
We can denote this candidate resident set as $R^*(e)$, a superset
of the true resident part set.
\[R(e) \subseteq R^*(e), R^*(e) = \cap_{c\in \downarrow e} R(c)\]

Finally, If the one-level downward adjacent set is known,
the existence and identity of an entity can be confirmed
using upward adjacencies.
Let $\uparrow$ denote the one-level upward adjacent set.
\[\uparrow e = e\{M^{d+1}\}, e \in M^d\]
Given a candidate downward set $(\downarrow e)^*$, we can
find $e$ as follows.
First, if $e$ exists on the same part, it will appear in
the upward sets of all the candidate downward entities.
\[\forall c\in (\downarrow e)^* : (e \in \uparrow c)\]
So, any upward set from a candidate downward entity
gives us a set of candidate entities $\{e\}^*$
to look through.
\[\{e\}^* = \uparrow c, c \in (\downarrow e)^*\]
The candidates can be tested, obviously, by comparing
their downward sets $\downarrow e$ with the candidate
downward set $(\downarrow e)^*$.

\section{Algorithm}

We now have enough proofs to write an algorithm that takes
as input the remote copy links of mesh vertices and
derives the remote copy links for the rest of the mesh.
Remote copy links are required to communicate identities
between parts, and we rightfully assume that having
remote copy information implies having resident part information
about an entity.

For every dimension starting from 1, the algorithm will establish
candidate resident part sets $R^*(e)$ for every entity on a part.
To every candidate part $p \in R^*(e)$ it sends a link proposal
consisting of the downward set $\downarrow e$ on the local part,
which on the remote part becomes a candidate downward set
$(\downarrow e)^*$.
It also sends the local identity of $e$, which
on the remote part becomes $e^*$, and a
remote copy link is made if $e$ is found on
the remote part.
Since all parts will perform this algorithm, it can be seen
that correct links will be established both ways between
copies.
At each dimension, full remote copy information is established
for all entities of that dimension, which is then used to
do the same for the next dimension up.

This algorithm will work correctly even in the case of ghosted
regions and pathological single-element parts,
but it only works for one part per process.
Making it work for multiple parts per process is left as
an exercise for the reader.

\pagebreak

\begin{algorithmic}
\Function{stitch}{$M, R(v)$ for $v \in M^0$}
\State Let $d_\text{max}$ be the highest entity dimension allowed
to have remote copies.
\State Let $p$ be the local part number
\For{$d$ from 1 up to and including $d_\text{max}$}
\For{$e \in M^d$}
\For{$b \in (R^*(e)\setminus p)$}
\State send $(e,\downarrow e)$ to $b$
\EndFor
\EndFor
\For{$(e^*,(\downarrow e)^*)$ received from $a$}
\State choose any $c \in (\downarrow e)^*$
\For{$e \in \uparrow c$}
\If{$\downarrow e = (\downarrow e)^*$}
\State add $(e^*,a)$ as remote copy of $e$
\EndIf
\EndFor
\EndFor
\EndFor
\EndFunction
\end{algorithmic}

\end{document}
